---
title: "Hierarchical (Multilevel) Models"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## The problem: estimating per-group means with very different sample sizes

Lots of data has a **grouping structure**: students nested in classes, patients in hospitals, products in stores, measurements in counties. We want to estimate something for *each* group, but some groups have lots of data and some have very little.

Two obvious strategies — both bad — are at the extremes:

- **Pool everyone together.** Assume every group is the same. Quick, but ignores real between-group variation.
- **Treat each group separately.** Estimate each group only from its own data. Honest in some sense, but groups with $n = 2$ get unreliable estimates.

**Hierarchical models** (also called **multilevel models**) sit between these extremes. They share information across groups *just enough* to stabilize estimates for the small groups, while letting the big groups speak for themselves. The amount of sharing is learned from the data.

This lecture follows the canonical Gelman & Hill radon example, then briefly looks at varying-slope models for voting patterns.

## Case study: radon in Minnesota counties

Radon is a radioactive gas that seeps from soil into basements; chronic exposure causes thousands of lung-cancer deaths per year in the US. The level varies from county to county (geology) and from home to home (basement vs. ground floor, sealing, etc.).

We have 919 radon measurements from 85 counties in Minnesota. We want to estimate the **typical log-radon level in each county** — useful for public-health prioritization.

```{r}
library(ggplot2)
library(dplyr)
library(lme4)

srrs2 <- read.table("data/srrs2.dat", header = TRUE, sep = ",",
                    stringsAsFactors = FALSE, strip.white = TRUE)
mn <- srrs2 %>% filter(state == "MN")
mn$radon     <- mn$activity
mn$log.radon <- log(ifelse(mn$radon == 0, 0.1, mn$radon))
mn$county    <- trimws(mn$county)

nrow(mn)
length(unique(mn$county))
```

The sample sizes per county are very unbalanced:

```{r}
county_n <- mn %>% count(county, sort = TRUE)
range(county_n$n)
head(county_n)
tail(county_n)
```

A few counties have dozens of measurements; many have only one or two. **Lac Qui Parle** has 2.

## Approach 1: Complete pooling

Assume every county has the same true log-radon level $\mu$. Just take the overall mean:

```{r}
mean(mn$log.radon)
sd(mn$log.radon)
```

Equivalently, `lm(log.radon ~ 1, data = mn)`. This is the **complete-pooling estimate**: every county gets the same value (about 1.22).

Pro: very stable estimate.
Con: ignores the real fact that radon levels do vary across counties.

## Approach 2: No pooling

Estimate each county independently — fit a separate mean per county:

```{r}
fit_no_pool <- lm(log.radon ~ county, data = mn)
```

`lm(y ~ county)` with a categorical predictor is exactly "fit a different intercept for each county and don't share anything across counties."

Let's pull out the per-county estimates and their standard errors:

```{r}
s <- summary(fit_no_pool)
coefs <- as.data.frame(s$coefficients)
coefs$county <- sub("^county", "", rownames(coefs))

subset_for_plot <- coefs %>%
  filter(county != "(Intercept)") %>%
  slice(c(2, 10, 22, 30, 36, 40, 50, 55, 60, 65))

ggplot(subset_for_plot, aes(x = reorder(county, Estimate), y = Estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = Estimate - `Std. Error`,
                    ymax = Estimate + `Std. Error`), width = 0.3) +
  coord_flip() +
  labs(x = NULL, y = "no-pooling estimate (log radon)",
       title = "Per-county intercepts with ±1 SE")
```

The error bars are huge for some counties — that's the no-pooling pathology. Let's look at Lac Qui Parle:

```{r}
mn %>% filter(county == "LAC QUI PARLE") %>% select(zip, log.radon)
```

Two data points. Whatever the no-pooling estimate gives, we shouldn't trust it much. Compare to **Anoka**, which has many:

```{r}
nrow(mn %>% filter(county == "ANOKA"))
```

In general, **larger error bars mean smaller sample size**. The no-pooling estimate doesn't share information across counties — every county is on its own.

## The intuition for partial pooling

For Lac Qui Parle, we have very little data — we should "borrow" some information from what other counties look like. The reasonable thing is a *weighted average* between:

- the county's own measurements (only 2 of them), and
- what counties in general look like (the overall mean).

If the county had 30 measurements, we'd weight its own data heavily. With 2, we should weight the overall mean heavily. With 0, we'd be fully relying on the overall mean.

The amount of pooling should depend on (a) the sample size in the group and (b) how variable group means are across the population. **Hierarchical models do this automatically.**

## The hierarchical model

We model the data as being generated in two stages.

Stage 1 (within county): each measurement is normal around its county's true mean:

$$y_i \sim N(\alpha_{j[i]}, \sigma_y^2),$$

where $j[i]$ is the county in which measurement $i$ was taken, and $\alpha_j$ is county $j$'s true log-radon level.

Stage 2 (across counties): the county means themselves are draws from a population:

$$\alpha_j \sim N(\mu_\alpha, \sigma_\alpha^2).$$

So we have **two levels** of randomness: variation between measurements within a county ($\sigma_y$), and variation between counties ($\sigma_\alpha$). The whole point of the model is that the $\alpha_j$ are not independent free parameters — they share a common distribution that the data informs.

The hyperparameters $\mu_\alpha$ and $\sigma_\alpha$ are estimated from the data, along with the $\alpha_j$ themselves.

### Why this gives partial pooling

The trichotomy can be read directly off the model:

- $\sigma_\alpha = 0$: every county has the same mean → **complete pooling**.
- $\sigma_\alpha = \infty$: county means are unrelated → **no pooling**.
- $0 < \sigma_\alpha < \infty$: county means are related but not identical → **partial pooling**.

Letting the data choose $\sigma_\alpha$ is what gives the right amount of sharing.

## Fitting it with `lmer`

The `lme4` package fits these models. The syntax `(1 | county)` reads as "a random intercept that varies by county":

```{r}
fit_partial <- lmer(log.radon ~ (1 | county), data = mn)
summary(fit_partial)
```

The key numbers:

- `Random effects` `county (Intercept)`: this is $\hat{\sigma}_\alpha$ — how much county means vary.
- `Random effects` `Residual`: this is $\hat{\sigma}_y$ — within-county spread.
- `Fixed effects` `(Intercept)`: this is $\hat{\mu}_\alpha$ — the average across counties.

### Vocabulary aside: random vs. fixed effects

The `lme4` output partitions parameters into "fixed effects" and "random effects":

- **Random effects**: parameters whose values are themselves *modelled* as draws from a distribution (the $\alpha_j$, drawn from $N(\mu_\alpha, \sigma_\alpha^2)$).
- **Fixed effects**: parameters that aren't drawn from a distribution — they're just point parameters to estimate (the population-level $\mu_\alpha$, and any non-grouped predictors).

The terminology is **inconsistent across textbooks and papers**, and pretty confusing. Just remember: random effects = "modelled as drawn from a distribution"; fixed effects = "everything else."

## Lac Qui Parle: the three approaches side by side

```{r}
ranef_county <- coef(fit_partial)$county
ranef_county["LAC QUI PARLE", , drop = FALSE]

mean(mn$log.radon[mn$county == "LAC QUI PARLE"])

mean(mn$log.radon)
```

- **No-pooling** estimate for Lac Qui Parle: the raw mean of its 2 data points.
- **Complete-pooling** estimate: the overall mean (~1.22).
- **Partial-pooling** estimate: somewhere between the two, pulled toward the overall mean because Lac Qui Parle has so few measurements.

That's the partial-pooling magic: information from the other 84 counties tells us that Minnesota counties don't have wildly different log-radon levels, so when a county gives us very little data, we shouldn't trust its raw mean too much.

For a county with many measurements (Anoka), partial pooling barely moves the estimate — the county's own data dominates.

## Adding predictors

The floor on which the measurement was taken matters: basement measurements (floor = 0) tend to be higher than ground-floor measurements (floor = 1).

```{r}
fit_floor <- lmer(log.radon ~ floor + (1 | county), data = mn)
summary(fit_floor)
```

The `floor` coefficient gives the average effect of moving up one floor on log-radon (holding county constant). It's negative — moving up reduces radon, as expected.

### Random slopes

We could also let the **effect of floor** vary by county — maybe in some counties basements matter more than in others (ceiling heights, ventilation, etc.):

```{r}
fit_random_slopes <- lmer(log.radon ~ floor + (floor | county), data = mn)
summary(fit_random_slopes)
```

The model is now

$$y_i \sim N(\alpha_{j[i]} + \beta_{j[i]} x_i,\ \sigma_y^2), \qquad
\begin{pmatrix} \alpha_j \\ \beta_j \end{pmatrix} \sim N\!\left(\begin{pmatrix}\mu_\alpha \\ \mu_\beta\end{pmatrix},\ \Sigma\right),$$

where $\Sigma$ allows the intercepts and slopes to be correlated across counties (do high-baseline counties have stronger floor effects?).

## Prediction

### New observation in a *known* county

Suppose we want to predict log-radon for a new ground-floor measurement in Lac Qui Parle. The fitted model gives us $\hat{\alpha}_{LQP}$ and $\hat\beta_{LQP}$; we plug into

$$y_{\text{new}} \sim N(\hat\alpha_{LQP} + \hat\beta_{LQP} \cdot 1,\ \hat\sigma_y^2).$$

```{r}
library(arm)  # for sigma.hat()
sig_y <- sigma.hat(fit_random_slopes)$sigma$data
coef_lqp <- as.numeric(coef(fit_random_slopes)$county["LAC QUI PARLE", ])

x_new <- 1
y_new <- rnorm(1000, mean = coef_lqp %*% c(1, x_new), sd = sig_y)
quantile(y_new, c(0.025, 0.975))
```

The 95% predictive interval for log-radon. Exponentiate to get the interval for raw radon.

### New observation in a *new* county

This is the interesting one. We don't have $\alpha$ and $\beta$ for the new county — but the model tells us how county-level parameters are distributed. We draw $(\alpha, \beta)$ from the **population distribution**, then draw a $y$ given those:

```{r}
library(MASS)  # for mvrnorm()

# Pull the estimated population-level parameters
vc <- VarCorr(fit_random_slopes)$county
Sigma <- matrix(vc, nrow = 2, dimnames = dimnames(vc))
mu    <- fixef(fit_random_slopes)

random_params <- mvrnorm(n = 1000, mu = mu, Sigma = Sigma)

x_new <- 1
y_new_unknown <- rnorm(1000,
                      mean = random_params %*% c(1, x_new),
                      sd = sig_y)
quantile(y_new_unknown, c(0.025, 0.975))
```

The interval is **wider** than for a known county — and it should be, because for a new county we have to integrate over our uncertainty about where its $\alpha$ and $\beta$ sit.

This kind of prediction is one of the big practical benefits of hierarchical models: you get principled uncertainty quantification for groups you've never seen.

## Another case study: voting patterns across US states

This is the running example in Gelman's *Red State, Blue State, Rich State, Poor State* (2009). The puzzle: rich *states* tend to vote Democratic, but rich *voters* tend to vote Republican. How can both be true?

Let $y_i = 1$ if person $i$ voted Republican, $x_i$ be income (rescaled), and $s[i]$ be their state.

A first ("baseline") model:

$$P(y_i = 1) = \operatorname{logit}^{-1}(\alpha_{s[i]} + \beta x_i).$$

State intercepts $\alpha_{s[i]}$ vary, but the income effect $\beta$ is the same in every state. This is a hierarchical logistic regression — fit as

```r
glmer(bush ~ income + (1 | state), data = polls, family = binomial)
```

A better model lets the income slope vary too:

$$P(y_i = 1) = \operatorname{logit}^{-1}(\alpha_{s[i]} + \beta_{s[i]} x_i).$$

```r
glmer(bush ~ income + (income | state), data = polls, family = binomial)
```

The fitted $\beta_{s}$'s differ substantially: **in poor states (Mississippi), income strongly predicts voting Republican. In rich states (Connecticut), income barely matters or even reverses.** That state-level heterogeneity is what dissolves the rich-state/rich-voter paradox.

This is a place where you really need a varying-slope hierarchical model — a plain logistic regression with state fixed effects would miss the substantive finding entirely.

## The Bayesian connection

Everything above can be reformulated as Bayesian inference:

- We put priors on the hyperparameters $\mu_\alpha,\ \mu_\beta,\ \sigma_\alpha,\ \sigma_\beta,\ \rho$.
- We get posterior distributions on everything — including the per-group $\alpha_j$ and $\beta_j$.
- The hierarchical structure is part of the prior: it says "the $\alpha_j$ are drawn from a common population," which is exactly what produces partial pooling.

Tools like Stan, `rstanarm`, and `brms` fit these models fully Bayesianly. For the radon example:

```r
library(rstanarm)
stan_lmer(log.radon ~ floor + (floor | county), data = mn)
```

returns posterior samples for every parameter, including each county's $\alpha_j$. Useful when you want full uncertainty quantification (e.g. for the rare-county-prediction case above) and when you want to use **weakly informative priors** to stabilize the model further on small data.

## A useful mindset

For grouped data, the spectrum is:

| | $\sigma_\alpha$ | What it says |
|---|---|---|
| Complete pooling | $0$ | "Groups are all the same" |
| Partial pooling | $0 < \sigma_\alpha < \infty$ | "Groups differ — but in a structured way" |
| No pooling | $\infty$ | "Groups are unrelated" |

The complete-pooling and no-pooling extremes are **special cases** of the hierarchical model, with $\sigma_\alpha$ forced to a particular value. Letting $\sigma_\alpha$ be estimated from the data is almost always better than picking an extreme. The intuition: if there's no group-level variation, the data will tell you so ($\hat\sigma_\alpha \approx 0$, partial pooling collapses to complete pooling). If groups are genuinely unrelated, the data tells you that too ($\hat\sigma_\alpha$ large, partial pooling collapses to no pooling). You don't have to choose.

Wherever you see grouped data — students in classrooms, patients across hospitals, products across stores, repeated measurements per subject — partial pooling is a strong default.

## References

- Gelman & Hill (2007), *Data Analysis Using Regression and Multilevel/Hierarchical Models*, Ch. 12 — the canonical textbook treatment.
- Gelman (2009), *Red State, Blue State, Rich State, Poor State* — the voting case study.
- Shalizi, *Advanced Data Analysis from an Elementary Point of View*, Ch. 21–24.
